---
output: html_document
editor_options: 
  chunk_output_type: console
---
```{r include=FALSE}
library(tidyverse)
library(rethinking)
library(conflicted)
library(tidytuesdayR)

```

### Let's take a look at the data.

```{r include= FALSE}

him <- tt_load(2025, week = 3)
exped <- him$exped_tidy
peaks <- him$peaks_tidy

# make any modifications to the dataset here.
peaks <- peaks %>% 
  mutate(
    pstatus = PSTATUS_FACTOR,
    himal_range  = HIMAL_FACTOR,
    peak_height = HEIGHTM,
    open = OPEN
    
  )

exped <- exped %>% 
  rename(route1 = ROUTE1,
         route2 = ROUTE2,
         route3 = ROUTE3,
         route4 = ROUTE4,
         )

exped <- exped %>% 
  rename(
    success1 = SUCCESS1,
    success2 = SUCCESS2,
    success3 = SUCCESS3,
    success4 = SUCCESS4
  )

exped <- exped %>% mutate(
  hired_personnel = !NOHIRED
)
```

#### What is the distribution of climbing status across different mountains ranges?
```{r}
peaks %>% 
  group_by(pstatus, himal_range) %>% 
  summarize(count = n()) %>% 
  pivot_wider(
    names_from = pstatus,
    values_from  = count
  ) %>% 
  mutate(
    total = Climbed + Unclimbed,
    climbed_prop = Climbed/total,
    himal_range = fct_reorder(himal_range, climbed_prop)
  ) %>% 
  arrange(desc(climbed_prop)) %>% 
  ggplot(aes(himal_range, climbed_prop))+
  geom_col()+
  coord_flip()+
  labs(x = "Mountain Ranges",
       y = "Climbed/Total Expeditions",
       title = "Distribution of climbing status across different mountain ranges.")+
  theme_minimal()

```

#### Which mountain range has the highest average peak height?

```{r}

peaks %>% 
  group_by(himal_range) %>% 
  summarize(
    avg_peak_height = mean(peak_height)
  )
```

#### Distribution of peak heights for peaks that are closed vs those that are open.

```{r}
peaks %>% 
 group_by(open) %>% 
 ggplot() +
  geom_density(aes(peak_height, col = open))+
  labs(
    title = "Distribution of peak heights by the open/closed status",
    x = "Peak height",
    y  = "Density"
  )+
  theme_minimal()
```

#### Which climbing routes have the highest success rates?

```{r}

total <- length(exped$success1)

#For route 1.

#Another way I may have learned just now.
success_rate1 <- exped %>% 
  summarize(sum(success1/n())) %>% 
  pull()

# For route 2:

success_rate2 <- exped %>% filter(!is.na(route2)) %>% 
  summarize(sum(success2)/n()) %>% 
  pull()

success_rate2

# There must be a better way to do this but for now this is it.

```

#### How does the use of oxygen affect summit success rates?
```{r}
exped %>% count(O2USED, O2NONE)
```

Common logical sense would imply !O2USED = O2NONE. But, looking at the counts, there are 10 observations where both O2USED and O2NONE is false. This doesn's make any sense so I removing these rows.
```{r}
oxy <- exped %>% filter(O2USED == TRUE | O2NONE == TRUE) %>% 
  mutate(o2used = O2USED)

```
We may have to make a little contingency table. So, I want the success rate by oxygen status i.e. success rate when oxygen == 1 and success rate when oxygen  == 0.

```{r}
# This is for route 1.
oxy_success <- oxy %>% filter(o2used == TRUE) %>% 
  summarise(sum(success1)/n()) %>% pull()

nooxy_success <- oxy %>% filter(o2used == FALSE) %>% 
  summarise(sum(success1)/n()) %>% pull()

tibble(category = c("Oxygen Used", "Oxygen Not Used")
       ,value = c(oxy_success, nooxy_success)) %>% 
  ggplot(aes(category, value))+
  geom_col(fill = "skyblue", col = 1, width = 0.5) +
  geom_label(aes(label = round(value,2)), vjust = -0.5) + 
  theme_minimal()+
  ylim(0, 1) +
  labs(y = "Success Rates",
       x = "Status of Oxygen Usage",
       title = "Difference in success rates of mountain climbinng by oxygen usage.")


```

#### Are expeditions with no hired personnel associated with higher or lower death rates?
To calculate death rates, first total = number of members + number of hired members. And total deaths = member deaths + hired personnel deaths.
```{r}
exped <- exped %>% mutate(
  total_members = TOTMEMBERS + TOTHIRED,
  total_deaths = MDEATHS + HDEATHS,
  prop_deaths = total_deaths/ total_members
)

exped %>% group_by(hired_personnel) %>% 
  summarise(deaths = sum(total_deaths))

exped %>% group_by(hired_personnel) %>% 
  summarise(sum(total_deaths)/n())

```
Expeditions with hired personnel are associated with higher deaths. The mechanism and the cause of this can be manifold.
 